---
title: "Gam"
author: "Jack B"
date: "2024-06-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load packages
```{r}
library(mgcv)
library(tidyverse)
library(purrr)
library(data.table)
library(terra)
library(sjPlot)
require(sf)
require(sp)
require(raster)
require(gdistance)
require(rgeos)
library(rmapshaper)
library(gratia)
library(viridis)
library(spbabel)
```

Read in data frame if need to
```{r}
n_fish = fread(here::here("jack-help", "gam", "n_fish.csv"), header = TRUE) 
```

Filter relevant columns
```{r}
n_fish <- n_fish %>% 
  dplyr::select(new_receiver_id, longitude, 
                latitude, study_year, origin, 
                year, week, new_season, month, max_fish, number_of_fish)

n_fish$origin <- as.factor(n_fish$origin)
```

Filter data to just spring
```{r}
n_fish_spring <- n_fish %>% filter(new_season == 'spring')
```

Convert data frame to UTM
```{r}
#Convert to UTM
n_fish_spring_utm <- n_fish_spring %>%
  st_as_sf(., coords=c("longitude", "latitude"), crs=4326) %>% 
  st_transform(32633) %>% 
  as(., "Spatial") %>% 
  as_tibble() %>% 
  dplyr::rename(x=coords.x1, y=coords.x2)
```


1. Running GAM with tp sline with a bivriate spatial variable (runs fine)

```{r}
mod1 <- gam(number_of_fish ~ origin + s(x, y, by = origin, k = 40) + 
              s(study_year, k = 3) + origin, offset = log(max_fish),
            data = n_fish_spring_utm, 
            family=nb(link = "log"),
            method = "REML")
```

Model check
```{r}
# gam.check(mod1) #Not great, but going with this for now
appraise(mod1)
```

Model output
```{r}
summary(mod1)
```

Plot model
```{r}
gratia::draw(mod1, contour = FALSE, scales = 'fixed', select = c(1,2))
```

2. Using a 'soap film smoother' to bound the GAM

Creating soap film smoother
```{r}
shape <- st_read(here::here("jack-help", 
                            "gam", 
                            "shapefile_new", 
                            "siljan_new.shp")) %>% 
  st_transform(32633) %>% 
  as(., "Spatial") %>% 
  sp::geometry(.)
plot(shape, col = "blue")
```

Simplify shape
```{r}
shape_simp <- ms_simplify(shape, keep = 0.01, keep_shapes = FALSE, explode = TRUE)
plot(shape_simp, col = "blue")
```

Creat knots for soap film smoother
```{r}
shape.xy.aut <- sptable(shape_simp)

shape.xy.aut <- shape.xy.aut %>%
  dplyr::rename(x = x_, y = y_)

#Make knots from the geographical extent of the observations
N <- floor((abs((max(n_fish_spring_utm$x)-min(n_fish_spring_utm$x)))/1000)) #every 1000 m
gx <- seq(min(n_fish_spring_utm$x), max(n_fish_spring_utm$x), length.out = N)
gy <- seq(min(n_fish_spring_utm$y), max(n_fish_spring_utm$y), length.out = N)
gp <- expand.grid(gx, gy)
names(gp) <- c("x","y")
plot(gp$x, gp$y)

#The GAM needs the border coordinates as a list of lists,
#where each list describes one border segment or island:
coords <- shape.xy.aut %>% dplyr::select(x,y,branch_)
names(coords) <- c("x", "y", "branch_")
borderlist <- split(coords, coords$branch_)
names(borderlist)

border.aut <- lapply(borderlist, `[`, c(1,2))
nr <- seq(1,3)
border.aut <- lapply(nr, function(n) as.list.data.frame(border.aut[[n]]))

#Make knots
lake_grid <- st_as_sf(shape_simp) %>%
  st_make_grid(cellsize = 1000, square = TRUE, what = "centers") %>%
  st_as_sf() 

st_geometry(lake_grid) <- "geometry"

#remove knots that fall outside the boundry
lake_intesects <- st_intersection(st_as_sf(shape_simp), lake_grid)

lake_knots <- lake_intesects %>%
  mutate(
    lon = st_coordinates(.)[,"X"],
    lat = st_coordinates(.)[,"Y"]
  ) %>%
  st_drop_geometry() %>%
  as.data.frame() %>%
  dplyr::select(lon, lat)

#And then check that the border and knots are in order with the soap_check function
#devtools::install_github("dill/soapcheckr")
library(soapcheckr)

#vignette("how_to_use_soapcheckr")

lake_knots <- lake_knots %>% dplyr::mutate(x = lon,
                                           y = lat) %>% dplyr::select(x,y)

#check knots
soap_check(border.aut, knots = lake_knots)
```

identify knots that fall outside boundry
```{r}
crunch_ind <- autocruncher(border.aut, lake_knots)
crunch_ind
# remove knots that are problematic
lake_knots <- lake_knots[-crunch_ind, ]
soap_check(border.aut, knots = lake_knots)
```

Gam with a soap film smoother
```{r}
mod2 <- gam(number_of_fish ~ s(x, y, by = origin, k = 40, bs = "so", xt = list(bnd = border.aut, nmax=1500)) + s(study_year, k = 3) + origin, offset=log(max_fish),
            data = n_fish_spring_utm, 
            family=nb(link = 'log'),
            method = "REML", 
            knots = lake_knots)
```

Model checks
```{r}
gam.check(mod2) #Not great
```

Model summary
```{r}
summary(mod2)
```

Plot model - does not work (Error in `mutate()`:ℹ In argument: `.loop = rep(seq_along(pts), each = pts)`.Caused by error:! `.loop` must be size 536 or 1, not 1446.)
```{r}
gratia::draw(mod2, contour = FALSE, scales = 'fixed', select = c(1,2))
```

Plot model
```{r}
plot(mod2)
```

